---
title: "Design Documentation"
vignette: >
  %\VignetteIndexEntry{Design Documentation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
brand: "../inst/brand/_brand.yml"
format:
  html:
    theme:
      - brand
      - ../inst/brand/colors.scss
      - ../inst/brand/mermaid.scss
---

```{r}
#| label: setup
#| include: false
# Function stubs are scaffolding — not run. The Mermaid diagrams are rendered
# natively by Quarto, independent of these R chunks.
knitr::opts_chunk$set(eval = FALSE)

# Get theme colors (semantic is first layer, named is under $palette)
# Check both directory levels depending on where we are bulding vignette from
brand_path <- if (file.exists("inst/brand/_brand.yml")) {
  "inst/brand/_brand.yml"
} else {
  "../inst/brand/_brand.yml"
}
colors <- yaml::read_yaml(brand_path)$color
```


# Introduction

This vignette is the design plan and blueprint for the package.
It is organized into multiple developmental **arms** — the load-bearing parts of the project.
Each arm is drawn as a Mermaid flowchart box listing its functions, sitting right above the stubs it describes.

By using this document as a developmental *blueprint*, we can quickly see how the package was put together and the reasoning.

The set-up is that for each family of functions in the arm of the project, we are setting up a function stubs that describe the signature and general purpose, without having it completed. This helps to make sure the needed components for this project are available.

The stubs are aggregated in a Mermaid diagram to serve as a visual map of progress. If the content is [blue]{.blue}, it currently is implemented, and if it is [orange]{.orange}, it is planned but not yet implemented.

## Goal

`{gram}` explores, annotates, and presents cardiac electrophysiology studies.
`{EGM}` owns vendor import and WFDB-compatible signal/annotation I/O. `{gram}`
starts after normalization to WFDB and does not duplicate the raw signal.

## Arms of the project

1. Data Backend
1. Visualization Engine
1. Annotation Interaction
1. Mark-up and Presentation

```{mermaid}
%%| label: project-arms

%%| uses mermaid.scss preferences for the boxes so the colors can be branded 
flowchart LR
    EGM["EGM + WFDB<br/>vendor import<br/>range I/O"]:::implemented
    Backend["Data Backend<br/>study handle<br/>overview cache"]:::implemented
    Viz["Visualization Engine<br/>uPlot explorer"]:::planned
    Annot["Annotation Interaction<br/>schema; query; edit"]:::planned
    Bookmarks["Bookmarks<br/>saved view state"]:::planned
    Presentation["Mark-up and Presentation<br/>tracing grammar; anime.js; export"]:::planned

    EGM --> Backend
    Backend --> Viz
    Backend --> Annot
    Annot --> Viz
    Viz --> Bookmarks
    Bookmarks --> Presentation
    Annot --> Presentation
```

# Data Backend

```{mermaid}
%%| label: data-backend-map
flowchart LR
    EGMIO["EGM I/O<br/>read_signal()<br/>read_annotation()<br/>write_annotation()"]:::implemented
    GramBackend["{gram} data backend<br/>study_cache()<br/>build_overview()<br/>read_viewport()"]:::implemented

    EGMIO --> GramBackend
```

## Study handle

The study cache or handle contains paths, header metadata, record fingerprint, cache manifest,
and annotation-layer metadata. 
It is referential only, and introduces an `S7` class called `StudyCache`.

```{r}
# Read header and locate sidecars; do not read the signal.
study_cache <- function(path, cache_dir = NULL, annotators = NULL, ...) {
  # implemented in R/cache.R
}

# Stream the WFDB record in fixed-size chunks and build derived tiers.
build_overview <- function(
  cache,
  chunk_seconds = 60,
  format = c("parquet", "rds"),
  rebuild = FALSE
) {
  # implemented in R/overview.R
}

# Return only the samples needed for a viewport; `window` is in samples.
read_viewport <- function(
  cache,
  window,
  channels = NULL,
  width_px = NULL,
  resolution = c("auto", "raw", "overview")
) {
  # implemented in R/overview.R
}
```

Files gram writes beside a record are named `<stem>.gram.*`, so one rule keeps
them out of annotator discovery and the package name says where they came
from. `<stem>.gram.json` is a manifest shared by every part of gram: each
writer reads it, replaces only its own section, and writes the whole file back
atomically. The cache owns the `cache` section; bookmarks and the annotation
sidecar will own sections of their own, and a rebuild can never clobber them.

Requirements:

- Raw `.dat`/`.hea` files remain canonical and unmodified.
- Raw windows come from `EGM::read_signal(begin, end, channels)`.
- Chunk size bounds peak R memory during cache construction.
- Cache is regenerable, versioned, and invalidated by the record fingerprint.
- Default cache location is user-writable; read-only study directories must work.
- Interrupted builds are resumable or safely discarded; publish cache atomically.

## Overview pyramid

Initial reduction: **min/max per channel per bucket**, retaining each extremum's
sample index and emitting the pair in time order. Advantages: narrow spikes are
not averaged away; levels can be built by merging adjacent buckets; one streaming
pass with bounded memory.

- Geometric bucket sizes, e.g. `64, 256, 1024, ...` samples.
- Choose the finest tier that returns at most about 2--4 points/pixel/channel.
- Use raw signal once the visible range is already near screen resolution.
- One table, all tiers stacked, in `<stem>.gram.parquet` (rds on request):
  `level`, `start`, then `ch<i>.min`, `ch<i>.min_at`, `ch<i>.max`,
  `ch<i>.max_at` per channel position. Parquet reads a subset of channels
  without touching the rest; rds cannot.
- The `cache` section of the manifest: version, algorithm, record fingerprint,
  table file and format, units, sampling frequency, sample count, channels,
  bucket sizes, and build time. Its presence is the completion marker, and a
  fingerprint that no longer matches the record reads as not built.
- Never silently use an overview tier for measurement or annotation snapping.

`LTTB` remains a benchmark alternative, not the first implementation. Clinical
acceptance tests should compare reductions on narrow His/stim artifacts and noisy
intracardiac channels.

## Viewport return contract

- Raw: one shared `sample` array plus aligned channel arrays.
- Overview: one `sample`/`value` pair per channel; extrema differ by
  channel. Collapse a min/max pair when both refer to the same sample.
- `time` is never stored or returned; it is `sample / sample_rate`.
- Sample representation must remain exact for the supported maximum study size.
- `resolution`: `"raw"` or the cache bucket size.
- `request_id`: used to reject stale asynchronous responses.

# Visualization Engine

```{mermaid}
%%| label: visualization-map
flowchart LR
    RangeReads["EGM range reads<br/>viewport source"]:::implemented
    UplotRenderer["uPlot renderer<br/>synced channels<br/>pan; zoom; cursor"]:::planned
    StudyExplorer["Study explorer<br/>explore_study()<br/>view_segment()"]:::planned

    RangeReads --> UplotRenderer
    UplotRenderer --> StudyExplorer
```

uPlot is the canvas renderer, not the data backend. Its numeric x-scale,
`setScale()`, `setData()`, hooks, and chart synchronization fit this use case;
its input must remain viewport-sized. See the
[uPlot API documentation](https://github.com/leeoniya/uPlot/blob/master/docs/README.md).

```{r}
# Server-backed explorer: R services viewport and annotation requests.
explore_study <- function(study, channels = NULL, annotators = NULL, ...) {
  # TODO
}

# Standalone widget for an already-materialized window; no dynamic disk reads.
view_segment <- function(x, annotations = NULL, ...) {
  # TODO
}
```

Design:

- Shiny controller for the MWE; a static htmlwidget cannot request new R reads.
- One uPlot per visible channel, synchronized on x-range and cursor.
- Separate charts permit independent y-gain and channel-specific min/max x values.
- One controller issues a single data request for all panels.
- Keep the old tier visible while the next tier loads; discard stale responses.
- Debounce zoom/pan requests; later prefetch a small margin in the pan direction.
- Overview navigator always shows the study extent and current viewport.
- Wheel/pinch zoom around cursor; drag pan; keyboard next/previous window.
- Channel order, visibility, gain, grid, and polarity are view state.
- Annotations render in a plugin layer, not as dense uPlot signal series.

The canonical interactive control is visible duration. Conventional sweep-speed
presets map to that duration. Exact `mm/s` is only meaningful after screen
calibration; print output can use physical dimensions exactly.

# Annotation Interaction

```{mermaid}
%%| label: annotation-map
flowchart LR
    AnnotIO["EGM annotation I/O<br/>annotation_table()<br/>read_annotation()<br/>write_annotation()"]:::implemented
    Schema["Annotation schema<br/>archetypes<br/>channel scope<br/>snap rules"]:::planned
    Review["Review and navigate<br/>query; jump<br/>add; move; save"]:::planned
    Bookmarks["Bookmarks<br/>saved view state"]:::planned

    AnnotIO --> Schema
    Schema --> Review
    Review --> Bookmarks
```

## Logical model

The UI depends on a richer logical model than a hard-coded list of QRS symbols.
The model is adapted to/from `{EGM}` annotation tables.

| Field | Purpose |
|---|---|
| `id` | Stable identity across edits |
| `code` | Schema-defined type: `QRS`, `A`, `H`, `V`, `stim`, ... |
| `sample`, `end_sample` | Point or span in canonical coordinates |
| `channel` | Header channel id; missing/zero means study-wide |
| `source` | Imported, algorithm, user, or named annotator |
| `group_id` | Relate pulses/deflections to an episode or beat |
| `attributes` | Typed schema-defined values |

Annotation schemas are user-supplied and versioned. Each archetype defines:
code, label, point/span/group geometry, channel scope, style, allowed attributes,
and optional snap rule (`none`, peak, trough, maximum absolute slope).

Programmed stimulation should be a span/group with child stimulus marks and
attributes such as site and coupling intervals. This supports useful filtering;
hundreds of unrelated ticks do not.

```{r}
annotation_schema <- function(definition) {
  # TODO: validate and register archetypes
}

load_annotations <- function(study, annotator, schema = NULL) {
  # TODO
}

query_annotations <- function(x, window = NULL, channel = NULL,
                              code = NULL, source = NULL, ...) {
  # TODO
}

save_annotations <- function(x, annotator, overwrite = FALSE) {
  # TODO: write a new EGM/WFDB annotator atomically
}
```

Interaction requirements:

- Filter by type, channel, source, group attributes, tags, and time range.
- Jump next/previous; select from a result list; show event density when zoomed out.
- Add, select, drag, delete, and reclassify; channel inferred from clicked panel.
- Snapping reads raw signal around the candidate sample, never an overview tier.
- Optimistic browser movement; R confirms canonical sample/channel state.
- Undo/redo within a session; unsaved-change indicator; explicit save.
- Imported/detector annotator is immutable by default. Save review as a new name.
- Reopening the saved annotator reproduces the edited sample and channel exactly.

WFDB remains the interoperable exchange layer. If stable ids, spans, relations,
or audit metadata cannot round-trip through WFDB fields, store them in a versioned
companion sidecar rather than silently packing semantics into `aux`. Finalize that
mapping only after the intracardiac archetypes are supplied.

## Bookmarks

A bookmark is a **saved view**, not copied signal data: record fingerprint,
sample range, channel order, gains, visible annotation layers, label, notes, and
tags. It appears in navigation like an event and is the input to `tracing()`.

```{r}
bookmark_view <- function(study, window, label = NULL, state = NULL, ...) {
  # TODO
}

list_bookmarks <- function(study, filter = NULL) {
  # TODO
}
```

# Mark-up and Presentation

```{mermaid}
%%| label: presentation-map
flowchart LR
    TracingGrammar["R tracing grammar<br/>tracing(); caliper()<br/>arrow(); emphasize(); reveal()"]:::planned
    SvgCompiler["SVG compiler<br/>one coordinate space<br/>explicit final state"]:::planned
    AnimeJs["anime.js v4<br/>timeline<br/>interpolation; playback"]:::planned
    Exports["Exports<br/>SVG; PDF<br/>HTML; later GIF/video"]:::planned

    TracingGrammar --> SvgCompiler
    SvgCompiler --> AnimeJs
    SvgCompiler --> Exports
    AnimeJs --> Exports
```

Exploration and presentation use different renderers. uPlot/canvas prioritizes
latency; short selected segments can be rendered as SVG for precise print and
cross-channel markup. Both consume the same samples, annotations, and bookmark
state.

## Grammar

Workspace for what to design for grammar. Series of examples to organize.

- Add H to mark His
- Add point on specific channel (snap to peak)
- dV/dt marker
- arrows from one channel to another (highlight activation sequence)
- vertical square or rectangle to highlight region
- vertical stripe through entire plot

## Animation engine: anime.js

**Decision: use anime.js v4.** It is a good fit for this layer:

- MIT licensed and small enough to vendor with the htmlwidget; no CDN required.
- Animates SVG attributes, DOM properties, and JavaScript values.
- [`createTimeline()`](https://animejs.com/documentation/timeline/) supports
  ordered and concurrent steps, labels, timers, callbacks, and nested timelines.
- Playback methods provide play, pause, restart, reverse, and seek controls.
- [SVG helpers](https://animejs.com/documentation/svg/) support line drawing,
  motion paths, and later morphing.

Responsibility boundary:

- `gram_scene`: semantic objects, timing, and explicit final state.
- SVG compiler: traces, arrows, calipers, labels, and channel layout.
- anime.js: interpolation and timeline playback only.
- Exporter: SVG/PDF from the explicit final state; animated HTML from the same
  scene plus anime.js.

This boundary keeps the R grammar stable if the JavaScript runtime changes. It
also ensures a print still does not depend on replaying an animation to discover
where its elements finish.

| Tracing operation | anime.js compilation |
|---|---|
| reveal trace/arrow | SVG drawable from `0` to `1` |

Vendor one pinned anime.js v4 build under `inst/htmlwidgets/lib/`. Keep selectors,
timeline positions, and anime.js option names inside the compiler rather than in
the public R object.

```{r}
tracing <- function(x, ...) {
  # TODO: create from a bookmark or explicit study window; returns a tracing object
  # Creates a S7 Tracing object
}

# Example of annotation
add_marker <- function(x, from, to, label = NULL, ...) {
  # TODO
  # Generic for adding objects: arrows, dots, letters, etc
}

emphasize <- function(x, target, ...) {
  # TODO
}

render_tracing <- function(x, format = c("svg", "pdf"), ...) {
  # TODO
}

animate_tracing <- function(x, autoplay = TRUE, controls = TRUE, ...) {
  # TODO: compile the scene timeline to an anime.js-backed htmlwidget
}
```
