---
title: "Design Documentation"
vignette: >
  %\VignetteIndexEntry{Design Documentation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
# Function stubs are scaffolding — not run. The nomnoml chunks ARE run so the
# diagrams render; each is marked `#| eval: true` individually.
knitr::opts_chunk$set(eval = FALSE)
```

# Introduction

This vignette is the design plan and blueprint for the package. 
It is organized into multiple developmental **arms** — the load-bearing parts of the project. 
Each arm is drawn as a `{nomnoml}` class box listing its functions, sitting right above the stubs it describes.

By using this document as a developmental *blueprint*, we can quickly see how the package was put together and the reasoning. 

The set-up is that for each family of functions in the arm of the project, we are setting up a function stubs that describe the signature and general purpose, without having it completed. This helps to make sure the needed components for this project are available.

The stubs are aggregated in a `{nomnoml}` diagram to serve as a visual map of progress. If the content is [blue]{.blue}, it currently is implemented, and if it is [orange]{.orange}, it is planned but not yet implemented.

## Arms of the project

1. Data backend
1. Visualization
1. Annotation layer
1. Presentation


## The four arms

```{r}
#| eval: true
#| echo: false
nomnoml::nomnoml("
#direction: right
#.exists: fill=#e8f5e9
#.new: fill=#fff3e0
[<exists>Data arm|
  WFDB + EGM reader;
  windowed reads;
  overview cache]
[<new>Visualization arm|
  fast uPlot explorer;
  zoom + measure]
[<new>Annotation arm|
  intracardiac schema;
  review + navigate]
[<new>Presentation arm|
  scenes;
  animate + flatten]

[Data arm] -> [Visualization arm]
[Data arm] -> [Annotation arm]
[Annotation arm] -> [Visualization arm]
[Annotation arm] -> [Presentation arm]
[Visualization arm] -> [Presentation arm]
")
```

- **Data** is the substrate: WFDB files and the `{EGM}` reader that pulls them
  into R in pieces, because the studies are too large to load whole.
- **Visualization** is the fast explorer: preview the whole study, zoom freely,
  measure on the fly.
- **Annotation** is the hard, new arm: an intracardiac annotation schema built on
  WFDB, plus the review tools — and it's also how you navigate the study.
- **Presentation** is the zoomed-in, storytelling arm: annotate a segment, then
  optionally animate intervals to make a finding legible to a reader.

One concept worth pinning down because it shows up in three arms: **measuring.**
*Computing* an interval from saved marks is `measure()` (Annotation). *Reading
one off the screen* ad hoc is `egm_caliper()` (Visualization). *Showing one to an
audience* is `caliper()` (Presentation). Same idea, three jobs, kept separate.

## Arm 1 — Data

The substrate and how the package gets data out of it efficiently. The `{EGM}`
core already exists; the new work is a derived **overview cache** (a
multi-resolution min/max pyramid in Parquet) so the explorer can be fed
viewport-sized data at any zoom without re-reading the `.dat`.

```{r}
#| eval: true
#| echo: false
nomnoml::nomnoml("
#direction: right
#.exists: fill=#e8f5e9
#.new: fill=#fff3e0
[<exists>EGM core|
  read_wfdb();
  read_signal() windowed;
  annotation_table();
  read_annotation();
  write_annotation()]
[<new>Overview cache|
  build_overview();
  read_overview();
  overview_is_current();
  zoom_level()]
[EGM core] -> [Overview cache]
")
```

```{r}
#| label: data-arm
# --- exists (anchors) ---
# read_wfdb(); read_signal(); read_header(); annotation_table();
# read_annotation(); write_annotation(); read_bard()/read_muse()/read_prucka()

# Build a multi-resolution min/max pyramid for a record; cache as Parquet beside
# it. One streaming pass over the .dat. The engine behind instant zoom.
build_overview <- function(record) {
  # TODO
}

# Return ~viewport-width min/max pairs per channel for a time range, at the
# pyramid level matching the requested resolution. Range query over the cache.
read_overview <- function(record, time_frame) {
  # TODO
}

# Is the cache present and consistent with the current .dat?
overview_is_current <- function(record) {
  # TODO
}

# Pick the pyramid level: bucket size ~= visible_samples / width_px. (internal)
zoom_level <- function(time_frame, width_px) {
  # TODO
}
```

## Arm 2 — Visualization

The fast explorer. A uPlot-backed widget (one synced uPlot per channel) renders
the overview data and refines as you zoom. Plus an interactive caliper for
quick, throwaway measurements.

```{r}
#| eval: true
#| echo: false
nomnoml::nomnoml("
#.new: fill=#fff3e0
[<new>Visualization|
  egm_widget() - uPlot explorer;
  egm_widget_update() - refine on zoom;
  egm_caliper() - interactive measure]
[Overview cache] -> [Visualization]
")
```

```{r}
#| label: visualization-arm
# uPlot-backed htmlwidget: the fast, zoomable, multi-channel explorer. Renders
# aggregated overview data; one synced uPlot per channel, shared x-zoom.
egm_widget <- function(x) {
  # TODO
}

# Push a higher-resolution window into a live widget (Shiny proxy) — the
# "coarse now, crisp a beat later" refinement on zoom.
egm_widget_update <- function(proxy, time_frame) {
  # TODO
}

# Interactive on-screen caliper: two points -> read off dt and d-amplitude.
# Ephemeral; not persisted. (For saved intervals see measure() in the Annotation arm.)
egm_caliper <- function(x) {
  # TODO
}
```

## Arm 3 — Annotation

The complex arm, and the one that doesn't exist yet. Everything here hangs off a
schema: an intracardiac annotation vocabulary expressed in WFDB's fields. Get the
schema right and the rest (detection, review, navigation, measurement, ML export)
becomes composable. Detectors *propose* candidates; humans *curate* into a new
annotator layer; nothing overwrites source detections.

```{r}
#| eval: true
#| echo: false
nomnoml::nomnoml("
#direction: right
#.core: fill=#ffe0b2
#.new: fill=#fff3e0
[<core>Annotation schema|
  annotation_schema();
  intracardiac vocabulary;
  mapped to WFDB type / subtype / number / chan]
[<new>Detectors| annotator() - emit candidates]
[<new>Review & curate|
  review_annotations();
  accept / reject / move / add]
[<new>Navigate| egm_navigate() - step by event]
[<new>Measure| measure() - AH / HV / CL from marks]
[<new>Overlay| add_annotations(); add_intervals()]
[<new>ML export| export_training_set()]

[Annotation schema] -> [Detectors]
[Detectors] -> [Review & curate]
[Review & curate] -> [Navigate]
[Review & curate] -> [Measure]
[Review & curate] -> [Overlay]
[Review & curate] --> [ML export]
[Annotation schema] -> [Overlay]
")
```

```{r}
#| label: annotation-arm
# THE CRITICAL PATH. Define the intracardiac annotation vocabulary (His, A, V,
# stim, near/far-field, fractionation, double potentials, ...) and how each maps
# onto WFDB fields (type / subtype / number / chan). Nothing else in this arm is
# stable until this is settled.
annotation_schema <- function() {
  # TODO
}

# Template/registry entry for a detector that emits candidate annotations as a
# WFDB annotator layer (its own .ann extension). Extends the annotators family.
annotator <- function() {
  # TODO
}

# Interactive review across stacked annotator layers: accept / reject / move /
# add. Writes a curated annotator without touching the source detections.
review_annotations <- function(x) {
  # TODO
}

# Step through the study by annotation events (next/prev His, beat, flagged
# finding). Annotations are how you navigate a multi-hour record.
egm_navigate <- function(x, type) {
  # TODO
}

# Compute an interval (AH, HV, CL, VA, stim-to-x) from two persisted marks. The
# formal, reproducible measurement — returned as a computed view, not stored.
measure <- function(x, from, to) {
  # TODO
}

# Render annotation marks / measured intervals onto a ggm or scene.
add_annotations <- function(x) {
  # TODO
}
add_intervals <- function(x) {
  # TODO
}

# Assemble (signal window, curated label) pairs from a curated annotator for ML.
export_training_set <- function(record) {
  # TODO
}
```

## Arm 4 — Presentation

The zoomed-in, storytelling arm. A scene is a declarative spec over one window;
verbs decorate it (display a caliper, emphasize a channel, reveal an arrow, zoom
in). A storyboard is an ordered set of scenes. One spec, two compilers:
`animate()` for the dynamic/educational version, `flatten()` for the static,
low-tech, printable version.

```{r}
#| eval: true
#| echo: false
nomnoml::nomnoml("
#direction: right
#.new: fill=#fff3e0
[<new>Scene|
  scene();
  caliper() - display an interval;
  emphasize();
  reveal();
  zoom_to()]
[<new>Storyboard| storyboard()]
[<new>Compile|
  animate() - SVG + JS (dynamic);
  flatten() - pdf / pptx / png (static)]
[Scene] -> [Storyboard]
[Storyboard] -> [Compile]
")
```

```{r}
#| label: presentation-arm
# A declarative spec over one egm window: record, channels, time, gain, paper
# speed, theme.
scene <- function(x) {
  # TODO
}

# An ordered list of scenes forming a presentation.
storyboard <- function(...) {
  # TODO
}

# --- scene verbs (pipeable) ---
# Display a measured interval to the audience (storytelling form of measure()).
caliper <- function(x, ...) {
  # TODO
}
# Highlight a channel/beat, dim the rest.
emphasize <- function(x, ...) {
  # TODO
}
# Reveal an element (arrow, label) — a transition step.
reveal <- function(x, ...) {
  # TODO
}
# Zoom the scene to a sub-window — a transition step.
zoom_to <- function(x, ...) {
  # TODO
}

# --- compilers ---
# Dynamic: interactive SVG + JS timeline.
animate <- function(x) {
  # TODO
}
# Static: one frame per scene, no animation needed. Same spec as animate().
flatten <- function(x) {
  # TODO
}
```

## The app (integration)

The arms aren't a fifth arm — they're composed by a Shiny app, one module per
arm that needs a UI. The Data arm has no UI; it serves the others.

```{r}
#| eval: true
#| echo: false
nomnoml::nomnoml("
#direction: down
#.app: fill=#e3f2fd
[<app>App| launch_egm()]
[Explore module| mod_explore_ui(); mod_explore_server()]
[Annotate module| mod_annotate_ui(); mod_annotate_server()]
[Present module| mod_present_ui(); mod_present_server()]
[App] -> [Explore module]
[App] -> [Annotate module]
[App] -> [Present module]
")
```

```{r}
#| label: app-shell
# Launch the app: Explore / Annotate / Present over a record directory.
launch_egm <- function(record_dir) {
  # TODO
}

mod_explore_ui <- function(id) {}   ; mod_explore_server <- function(id) {}
mod_annotate_ui <- function(id) {}  ; mod_annotate_server <- function(id) {}
mod_present_ui <- function(id) {}   ; mod_present_server <- function(id) {}
```

## Status & open decisions

- **Critical path:** the `annotation_schema()` — the intracardiac vocabulary on
  WFDB fields. The Annotation arm can't stabilize until this is settled; it also
  shapes navigation, measurement, and ML export.
- **First build:** Data arm cache + Visualization arm widget as one vertical
  slice — open a ~1 GB study and drag-zoom smoothly. De-risks everything else.
- **Overview cache placement:** filed under Data (it *provides* data); the
  Visualization arm consumes it. Flag if you'd rather it sit under Visualization.
- **Cache backend:** Parquet via `arrow`, range queries via DuckDB. (`fst`
  rejected — R-only, format has broken across versions.)
- **Renderer:** uPlot (MIT, canvas); thin htmlwidget binding is ours to maintain.
- **Animation:** Web Animations API + d3 first (MIT, bundleable); GSAP via CDN
  only if path-morphing is needed — don't bundle GSAP in a CRAN package.
- **Open:** does the WFDB annotation schema stretch to dense intracardiac
  annotation (fractionation, double potentials, multichannel activation)? This is
  the question `annotation_schema()` has to answer.
```
